Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-16 MScDiplom_Lemenkova.tex
briefly described methods of stat analysis (Results), referenced Tables

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species (Fig. \ref{fig:2}), belonging to two orders
(Hydrocharitales and Najadales), four families (Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae), and 12 genera (Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera and Phyllospandix) \cite{Kuo89}\label{Kuo89}.
\begin{wrapfigure}{r}{0.4\textwidth}
\centering
	\includegraphics[scale=0.25]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world.\\ Source:\cite{Green03}\label{Green03} \label{fig:2}} 
\end{wrapfigure}
The distribution of the seagrasses is strongly influensed by several environmental factors, which
include climate (mostly, tropical and temperate areas), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutritient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are\textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\begin{wrapfigure}{r}{0.4\textwidth}
\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses.
	\label{fig:3}
Source: \url{http://www.ozcoasts.org.au}}
\end{wrapfigure}

@@

The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjasted for the marine
environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. WASI enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, which is a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.
The specific parameters chosen for the simulation of the environmengal conditions where seagrass
grow. The adaptations to life in salt sea water requires various phisical and chemical parameters
which include salinity, temperature of 17-25°C, light requirements with 10-20 percent on average, ranging
from 4.4percent  minimal up to 29 percent \cite{McKenzie09}\label{McKenzie09}, so that the zenith angle is taken as 35-
45° and reflection factor 0.0201. P.oceanica. 
The calculations are done for the spectrum 350-800 nm, covering the most important part of the RS spectrum: 1) Blue-green 0.45 - 0.5 μm; 2) Green 0.5 - 0.6 μm; 3) Red 0.6 - 0.7 μm; 4) Red-NIR 0.7 - 0.8 μm (Fig.\ref{fig:30}).

@@

In the statistical anlysis of the raw observed data pool of the values of spectral reflectances (about 400 single measurements for P.oceanica), we summarize their complexity by concentrating on some simple numerical characteristics that they possess, i.e. parameters. Examples are the mean and variance of a probability distribution of measurements dataset (Fig.). We divided the total data pool into datasets of 25 measurements for more convenience and ease of use, both for P.oceanica and carbonate sand. The mean of a probability distribution of spectral reflectance describes the average value of the random variable over all of its possible realizations (see for example, Fig. Appendix).  Conceptually, there are an infinite number of such realizations, therefore parameters are not known to us. However, in the statistical analysis of the observed data our goal is to estimate these parameters using a finite amount of information available to us. 
Thus, we collect a number of realizations (a sample of 25 measurements) and then estimate the unknown parameters using a statistical analysis (Appendices, see Tables ). \\
Finally we analysed 400 datasets made on two different data. 

@@

Probability thus expresses uncertainty about the possible values of the values of spectral reflectances  of P.oceanica.
Each time we observe the outcome of a random variable of spectral reflectance, we obtain an observation. 
Once observed, its value is known and hence it is no longer random. 
So, there is a distinction to be made between variables whose values are not yet observed (random variables) and those whose values have been observed (observations). 
We applied various methods of statistical probability analysis towards data pool, in order to analyse the distribution of values of spectral reflectances of P.oceanica (see Appendices, Tables). 
A probability distribution, a mathematical statement about the possible values that the random observations of spectral reflectance can take on, displays the relative frequency with which each possible value of spectral reflectance is observed.
hyperspectral radiometer Ramses: 350 measuremetn sets of P.oceanica reflectance for 14th Oct, 400
sets of P.oceanica reflectance for 15th Oct, 84 datasets for seawater reflectance with sediments, 105
Fig. 4.11. Spectral reflectance of water with sediments
Therefore, a statistical approach is necessary for the processing of such data. Statistical pre-processing
of large sets of serial data enables to generalyse data by using the most typical and predicted values of
The statistical calculators were mainly made by means of Gretle and SPSS. The Open Office was used for preliminary data view and preprocessing, and included following computations
at each data set (table 1): mean, median, Q1 and Q3, standard deviation, min and max values. The
statistical pre-processing and analysis were made shown the mean values of spectral reflectances,
seawater with and without sediment.

@@

the simplicity of plotting and the ease of scripting. (see Fig.15 and Appendix).
The graphs illustrating spectral signatures of various seafloor cover types display the mean values and
areas of quartiles (Q1, Q3, shaded areas on graphs), which cover most probable data distribution
(spectral reflectance) for each spectral band. In x-axis I displayed the areas of 400 – 950 nm in
spectra, where the measurements were done; the y-axis was adjusted for the better visualisation of the
P.oceanica spectra: as its spectral values are mostly located in the lower part of the spectra (usually no
more than 0.20 nm, except for borders), I did not extend the y-scale up to 100 percent, in order to
focus on the most necessary area of spectral values (Fig. 15 and Appendices). The mean values are
highligted using bold points, as I used these values for plotting the final graph of P.oceanica spectra in
Ligaria beach, Agia Pelagia.

@@

A statistical hypothesis test has been applied for the making decision and controlling the wealth of the
observational data of of the hyperspectral measurements of the water reflectances.
The received results are statistically significant if they are unlikely to have occurred by chance alone,
according to a pre-determined threshold probability, the significance level. Therefore, we applied
critical tests of significance to analyse the measured data according to their significant value.
Answering the first research question, we suggest the following statement.
Assuming that the null hypothesis, Hypothesis Ho, is true, which is “the spectral distinguishability of
the seagrass P.oceanica from other seafloor types is not changing with varying in-situ conditions,
Ho: μ1 =μ2 =μ3=...= μn.”, what is the probability of observing the values of spectral reflectance of
P.oceanica at various depths, for the test statistic, that is at least as extreme as the values of spectral
reflectance of P.oceanica at these depths, that was actually observed? The alternative Hypothesis Ha
Fig.15. Multiplot display of spectral reflectances of P.oceanica. Measurement sets 1-100, 15th
October. Bold red dots show the mean values within each dataset; vertical areas: quartiles Q1-Q2.
claims the opposite statement: “the spectral discernibility of the seagrass P.oceanica is distinctly
changing with varying in-situ conditions, i.e. increasing depth, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.
Two statistical approaches have been tested for confirmatory data analysis and hypothesis testing.
To analyse average values of spectral reflectance of P.oceanica by means of the measured sets we calculated 
mean values for the total set of emasurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the table 5 (Appendices). 
As a result, we have mean values of spectral reflectances of P.oceanica for extreme, average, median and both quartiles,
independent of the individual sets 1-350.

\paragraph{ANOVA-testing}
We applied ANOVA (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of P.oceanica to be more or less spectrally disticnt from other seafloor cover
types with changing depth. W are checking the statement Hypothesis Ho, is true, which is “the
spectral distinguishability of the seagrass P.oceanica from other seafloor types is not changing with
varying in-situ conditions, Ho: μ1 =μ2 =μ3=...= μn.” which gives the following outcome.
P(reject H0|H0 is valid)= P(X>c|p=)=.05
where c comes for critical values and p=probability.
The result P=.05 is, hence, very small, which makes the statement of Hypothesis Ho highly unlikely
(less than 1 in a 10 chance). The one-way ANOVA highlighted a significant difference between data
of spectral reflectances of P.oceanica at different depths, i.e. true is the opposite statement:
Hypothesis Ha=“the spectral discernibility of the seagrass P.oceanica is distinctly changing with
varying in-situ conditions, i.e. increasing depth, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.

@@

\subsubsection{Results of the remote sensing application}
The in situ spectral reflectance data of \textit{P.oceanica} and sand were used to model these seafloor cover
types as the different sensors would percept them: MODIS, ASTER, MERIS, SeaWiFS and CZCS.
These models simulate spectral views of the chosen sensors, how these sensors will "see" seagrass (Fig.\ref{fig:28}) and sand as
pixels, with accepted default atmosphere and water column effects (given by WASI software). In such
a way we defined an empirical upper limit to the discriminative potential of these sensors. To create
these sensor-specific reflectance spectra, spectral responses of each of these sensor were applied for
the spectral signature of seafloor cover types in full-resolution spectra.

@@

\begin{appendices}
\pagenumbering{roman}
\appendix
\appendixpage
\addappheadtotoc
\appendixname{\textbf{1:\\Results of the measurements of seafloor spectral reflectance}}
%\setcounter{figure}{0}

@@

\begin{figure}[h]
	\begin{center}
	\includegraphics[scale=0.20]{GNU-Stat-M-401-420.jpg}
	\caption{Fragment of the statistical analysis of the P.oceanica reflectance. Example of data set 401-420. 
		Visualisation of the measured data together with styatistical values: interquartile ranging, medians, means.}
	\label{fig:46}
	\end{center}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.25]{GNU-Radiance_Bezier.jpg}
	\caption{Radiance of the seawater with sediments. Measured in aquarium tank. Bezier Interpolation}
	\label{fig:47}
\end{figure}
\begin{table}[htbp]
	\rowcolors{1}{Ivory}{GhostWhite}
	\caption{Results of the \textit{ANOVA one-way} analysis: results of the \textit{single factor} (depth) testing of the radiance of\textbf{P.oceanica} at various depths: 0.5, 		1.5 and 2.5 meters}
	\begin{center}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline\hline
		\textbf{Source of Variation} & \textbf{SS} & \textbf{df} & \textbf{MS} & \textbf{F} & \textbf{P value} & \textbf{F crit} \\ \hline
		Between Groups & 373841.7048 & 2 & 186920.8524 & 407.85359 & 1.11677 & 3.01153 \\ \hline
		Within Groups & 261233.1668 & 570 & 458.3038014 & & & \\ \hline
		Total &  635074.8716 & 572 & & & & \\ \hline
	\end{tabular}
	\end{center} 
	P more than .05, which means that there is a significant difference in radiance of P.oceanica at three different depth (0.5, 1.5 and 2.5). 
	\label{tab:8}
\end{table}
\begin{table}[htbp]
	\rowcolors{1}{Ivory}{GhostWhite}
	\caption{Summary of the \textit{ANOVA one-way} analysis: results of the \textit{single factor} (depth) testing of the radiance of\textbf{P.oceanica} at various depths: 		0.5, 1.5 and 2.5 meters}
	\begin{center}
	\begin{tabular}{|c|c|c|c|c|}
		\hline\hline
		\textbf{Groups} & \textbf{Count} & \textbf{Sum} & \textbf{Average} & \textbf{Variance} \\ \hline\hline
		Column 1 & 191 &	760.126692 & 3.9797209 & 18.6948046 \\ \hline
		Column 2 & 191 &	4527.214218 & 23.7026922  & 251.4863978 \\ \hline
		Column 3 & 191 &	12465.19972 & 65.2628257 & 1104.7302015 \\ \hline
	\end{tabular}
	\end{center}
	P$>$0.05, which means that there is a significant difference in radiance of P.oceanica at three different depth (0.5, 1.5 and 2.5). 
 	\label{tab:9}
\end{table}
\pagebreak

\rowcolors{1}{GhostWhite}{LightCyan}
\begin{longtable}{|c|c|c|c|c|c|c|}
	\caption{Results of the statistical analysis of spectral reflectance of \textbf{P.oceanica}, sets 1-350). Wavelength step: 3 nm. Measured on Agia Pelagia beach, 15th October} \\
	  \multicolumn{1}{|l|}{\textbf{wl}} &
	   \multicolumn{1}{l|}{\textbf{mean}} & 
	   \multicolumn{1}{l|}{\textbf{Q1}} & 
	   \multicolumn{1}{l|}{\textbf{min}} & 
	   \multicolumn{1}{l|}{\textbf{Q3}} & 
	   \multicolumn{1}{l|}{\textbf{max}} & 
	   \multicolumn{1}{l|}{\textbf{median}} \\ \hline
	\endfirsthead

@@

\label{tab:10} 

@@

\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.25]{GNU-Irr-Splines.jpg}
		\caption{Irradiance of the seawater, measured in aquarium tank. Smooth Splines interpolation. Visualization in GNUplot­}
		\label{fig:48}
	\end{center}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.25]{GNU-Irr.jpg}
	\caption{Irradiance of the seawater, measured in aquarium tank. Smooth Bezier interpolation. Visualization in GNUplot}
	\label{fig:49}
\end{figure}
\pagebreak
\newpage
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.4]{G-gamma-rad.jpg}
		\caption{Frequency normality test against gamma distribution: radiance of the seawater, measured in aquarium tank. Visualization in Gretl.­}
		\label{fig:50}
	\end{center}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.4]{G-normal-dist.jpg}
	\caption{Frequency normality test against normal distribution: radiance of the seawater, measured in aquarium tank. Visualization in Gretl.­}
	\label{fig:51}
\end{figure}

@@

%%% Gretl start
\begin{table}[H]
	\rowcolors{1}{Thistle}{MintCream}
	\begin{center}
		\caption{Summary statistics of measurements set: 301-325. \\ Example of selected variables 12-22. Gretl.­}
		Summary Statistics, using the observations 301--325\\

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.38]{GNU-15.jpg}
		\caption{Multiplot of spectral reflectance of P.oceanica. Series 151-200. Shown midspread of the statistical quartiles Q1 and Q3 (vertical dashes) and mean value within the range (red bold dots). GNUplot visualization­}
		\label{fig:59}
	\end{center}
\end{figure}
\pagebreak

>>>>

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species (Fig. \ref{fig:2}), belonging to two orders
(Hydrocharitales and Najadales), four families (Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae), and 12 genera (Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera and Phyllospandix) \cite{Kuo89}\label{Kuo89}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world.\\ Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

The distribution of the seagrasses is strongly influensed by several environmental factors, which
include climate (mostly, tropical and temperate areas), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutritient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are\textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\begin{wrapfigure}{r}{0.4\textwidth}
\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

@@

The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjasted for the marine
environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. WASI enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, which is a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.
The specific parameters chosen for the simulation of the environmengal conditions where seagrass
grow. The adaptations to life in salt sea water requires various phisical and chemical parameters
which include salinity, temperature of 17-25°C, light requirements with 10-20 percent on average, ranging
from 4.4percent  minimal up to 29 percent \cite{McKenzie09}\label{McKenzie09}, so that the zenith angle is taken as 35-
45° and reflection factor 0.0201. P.oceanica. These values, simyulating the environment of the Mediterranean Sea, are fixed (Tab.\ref{tab:3}) on the WASI user-defined parameters.
The calculations are done for the spectrum 350-800 nm, covering the most important part of the RS spectrum: 1) Blue-green 0.45 - 0.5 μm; 2) Green 0.5 - 0.6 μm; 3) Red 0.6 - 0.7 μm; 4) Red-NIR 0.7 - 0.8 μm (Fig.\ref{fig:30}).

@@

hyperspectral radiometer Ramses: 350 measuremetn sets of P.oceanica reflectance for 14th Oct (Tab.\ref{tab:32}), 400
sets of P.oceanica reflectance for 15th Oct (Tab.\ref{tab:33}), 84 datasets for seawater reflectance with sediments, 105
A statistical approach is evidently necessary for the proper processing of such amounts of data. 
Statistical pre-processing of large sets of serial data enables to generalyse data by using the most typical and predicted values of
The statistical calculators were mainly made by means of Gretle and SPSS. 
The Open Office was used for preliminary data view and preprocessing, and included following computations
at each data set (for example, Tab.\ref{tab:31}) with summary outcome of mean, median, Q1 and Q3, standard deviation, min and max values (Tab.\ref{tab:4}). \\
The statistical pre-processing and analysis were made to display the mean values of spectral reflectances,
seawater with and without sediment (e.g. Tab.\ref{tab:2}).
In the statistical anlysis of the raw observed data pool of the values of spectral reflectances (about 400 single measurements for P.oceanica, e.g. Tab.\ref{tab:10}), we summarize their complexity by concentrating on some simple numerical characteristics that they possess, i.e. parameters. Examples are the mean and variance of a probability distribution of measurements dataset (for example, Tab.\ref{tab:16}). We divided the total data pool into datasets of 25 measurements for more convenience and ease of use, both for P.oceanica and carbonate sand. The mean of a probability distribution of spectral reflectance describes the average value of the random variable over all of its possible realizations (Appendices).  Conceptually, there are an infinite number of such realizations, therefore parameters are not known to us. However, in the statistical analysis of the observed data our goal is to estimate these parameters using a finite amount of information available to us: fieldwork observations. \\
Thus, we collect a number of realizations (a sample of 25 measurements for both sand and P.oceanica) and then estimate the statistical parameters (Appendix). \\
Finally we analysed abount 400 datasets totally (example of carbonate sand on Fig.\ref{tab:31}) made during different days. 

@@

Thus, we analysed the statistical distribution of the spectral reflectance values of seafloor cover types using common mathematical formulae, implmented in Gretl software (Tab.\ref{tab:19}). 
Their probability expresses uncertainty about the possible values of the spectral reflectances  of P.oceanica and sand, respectively.
There is a distinction to be made between variables whose values are not yet observed (random variables) and those whose values have been observed (observations). 
Each time we observe the outcome of a random variable of spectral reflectance, we obtain an observation, which is hence no longer random. 
We applied various methods of statistical analysis towards data pool (some examples of the selected tables are in the Appendix: Tab.\ref{tab:18}, Tab.\ref{tab:24}, Tab.\ref{tab:25}, Tab.\ref{tab:26}, Tab.\ref{tab:29}), in order to analyse the distribution of values of spectral reflectances of \textit{P.oceanica}. 
A probability distribution, a mathematical statement about the possible values that the random observations of spectral reflectance can take on, displays the relative frequency with which each possible value of spectral reflectance is observed. \\
The \textit{least absolute deviations (LAD)}, a popular optimization technique, was used to show the main trend of the distribution of spectral reflectance values along the spectra (Tab.\ref{tab:12}). Non-linear logistic anlysis is applied to specify and estimate a model of spectral distribution, in which the dependent variable (i.e. the value of spectral reflectance for each single observation) is not continuous, but discrete and independent for each case (Tab.\ref{tab:11}). \\
As the observations were made in a repeated way, we also analysed the data pool using autocorrelation as well: Tab.\ref{tab:13} and Tab.\ref{tab:14} for \textit{P.oceanica} and Tab.\ref{tab:28} for carbonate sand. The \textit{autocorrelation}, a cross-correlation of an observed value with itself, made on two different days (14th and 15th October),
 shows the similarity between observations as a function of the time separation between them. As can be seen on Fig.\ref{fig:52}, Fig.\ref{fig:53} and Fig.\ref{fig:60}, it is a useful tool for finding repeating patterns for a single value of observations, such as the presence of a periodic signal which has been buried under noise. The autocorrelation analysis has been performed using both SPSS and Gretl software. \\
The method of \textit{least squares} minimizes the sum of the squares of the errors made in solving every single observation made. We tried different approaches of the \textit{least squares} methods: Tab.\ref{tab:15}, Tab.\ref{tab:18}, Tab.\ref{tab:27}.\\
To focus on the relationship between a dependent variable (each single observation of the spectral reflectance) and independent variables (three various depths: 0.5, 1.5 and 2.5 meters) we used the \textit{regression analysis}, which enables modeling and analyzing values of the spectral reflectances at several depths: Tab.\ref{tab:5}, Tab.\ref{tab:6}, Tab.\ref{tab:7}. 
The summary of the values of spectral reflectance properties on various depths is presented on the Fig.\ref{fig:17}.
The \textit{k-means cluster analysis} has been used for partitioning observation sets into k clusters (Tab.\ref{tab:20}, Tab.\ref{tab:21}, Tab.\ref{tab:22}, Tab.\ref{tab:23}) in which each observation belongs to the cluster with the nearest mean.
It enables to highlight the main areas of the location of values of spectral reflectances along the spectra.\\
To estimate spectral density of a spectral reflectancfe measurements we used periodogram function (Tab.\ref{tab:30}), 
which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram of the series of observation highlights so-called “dead spots” of low power between the frequencies we wishe to
exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:43} ) 
and the frequencies we want to retain (the general profile that lies in the 450-850 nm, Fig.\ref{fig:45}, enlarged part of 500-660nm).\\

@@

the simplicity of plotting and the ease of scripting. (see Fig.\ref{fig:54}, Appendices).

The graphs illustrating spectral signatures of various seafloor cover types display the mean values and
areas of quartiles (Q1, Q3, shaded areas on graph, Fig.\ref{fig:55}), which cover most probable data distribution
(spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in
spectra, where the measurements were done; the y-axis was adjusted for the better visualisation of the
P.oceanica spectra: as its spectral values are mostly located in the lower part of the spectra (usually no
more than 0.20 nm, except for borders), we did not extend the y-scale up to 100 percent, in order to
focus on the most necessary area of spectral values (Fig.\ref{fig:35} and Appendices). The mean values are
highligted using bold red points, as we used these values for plotting the final graph of P.oceanica spectra in
Ligaria beach, Agia Pelagia.

@@

A statistical hypothesis test has been applied for the making decision and controlling the wealth of the
observational data of of the hyperspectral measurements of the water reflectances.
The received results are statistically significant if they are unlikely to have occurred by chance alone,
according to a pre-determined threshold probability, the significance level. Therefore, we applied
critical tests of significance to analyse the measured data according to their significant value.
Answering the first research question, we suggest the following statement.
Assuming that the null hypothesis, Hypothesis Ho, is true, which is “the spectral distinguishability of
the seagrass P.oceanica from other seafloor types is not changing with varying in-situ conditions,
Ho: μ1 =μ2 =μ3=...= μn.”, what is the probability of observing the values of spectral reflectance of
P.oceanica at various depths, for the test statistic, that is at least as extreme as the values of spectral
reflectance of P.oceanica at these depths, that was actually observed? The alternative Hypothesis Ha
Fig.15. Multiplot display of spectral reflectances of P.oceanica. Measurement sets 1-100, 15th
October. Bold red dots show the mean values within each dataset; vertical areas: quartiles Q1-Q2.
claims the opposite statement: “the spectral discernibility of the seagrass P.oceanica is distinctly
changing with varying in-situ conditions, i.e. increasing depth, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.
Two statistical approaches have been tested for confirmatory data analysis and hypothesis testing.
To analyse average values of spectral reflectance of P.oceanica by means of the measured sets we calculated 
mean values for the total set of emasurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (Appendices). 
As a result, we have mean values of spectral reflectances of P.oceanica for extreme, average, median and both quartiles,
independent of the individual sets 1-350.

\paragraph{ANOVA-testing}
We applied ANOVA (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of P.oceanica to be more or less spectrally disticnt from other seafloor cover
types with changing depth. W are checking the statement Hypothesis Ho, is true, which is “the
spectral distinguishability of the seagrass P.oceanica from other seafloor types is not changing with
varying in-situ conditions, Ho: μ1 =μ2 =μ3=...= μn.” which gives the following outcome.
P(reject H0|H0 is valid)= P(X$>$c|p=)=.05
where c comes for critical values and p=probability.
The result P=.05 is (Tab.\ref{tab:8}), hence, very small, which makes the statement of Hypothesis Ho highly unlikely
(less than 1 in a 10 chance). The one-way ANOVA highlighted a significant difference between data
of spectral reflectances of P.oceanica at different depths (Tab.\ref{tab:9}), i.e. true is the opposite statement:
Hypothesis Ha=“the spectral discernibility of the seagrass P.oceanica is distinctly changing with
varying in-situ conditions, i.e. increasing depth, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn“.

@@

\subsubsection{Results of the remote sensing application}
The in situ spectral reflectance data of \textit{P.oceanica} and sand were used to model these seafloor cover
types as the different sensors would percept them: MODIS, ASTER, MERIS, SeaWiFS and CZCS.\\
These sensors possess over various percepting characteristics which we briefly illustrated in a small summary table (Tab.\ref{tab:1})
These models simulate spectral views of the chosen sensors, how these sensors will "see" seagrass (Fig.\ref{fig:28}) and sand as
pixels, with accepted default atmosphere and water column effects (given by WASI software). In such
a way we defined an empirical upper limit to the discriminative potential of these sensors. To create
these sensor-specific reflectance spectra, spectral responses of each of these sensor were applied for
the spectral signature of seafloor cover types in full-resolution spectra.

@@

\begin{appendices}
\pagenumbering{roman}
\appendix \label{app}
\appendixpage
\addappheadtotoc
\appendixname{\textbf{1:\\Results of the measurements of seafloor spectral reflectance}}
%\setcounter{figure}{0}

@@

\begin{figure}[h]
	\begin{center}
	\includegraphics[scale=0.20]{GNU-Stat-M-401-420.jpg}
	\caption{Fragment of the statistical analysis of the P.oceanica reflectance. Example of data set 401-420. 
		Visualisation of the measured data together with styatistical values: interquartile ranging, medians, means.}
	\label{fig:46}
	\end{center}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.25]{GNU-Radiance_Bezier.jpg}
	\caption{Radiance of the seawater with sediments. Measured in aquarium tank. Bezier Interpolation}
	\label{fig:47}
\end{figure}

\begin{table}[htbp]
	\rowcolors{1}{Ivory}{GhostWhite}
	\caption{Results of the \textit{ANOVA one-way} analysis: results of the \textit{single factor} (depth) testing of the radiance of\textbf{P.oceanica} at various depths: 		0.5, 1.5 and 2.5 meters}
	\begin{center}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline\hline
		\textbf{Source of Variation} & \textbf{SS} & \textbf{df} & \textbf{MS} & \textbf{F} & \textbf{P value} & \textbf{F crit} \\ \hline
		Between Groups & 373841.7048 & 2 & 186920.8524 & 407.85359 & 1.11677 & 3.01153 \\ \hline
		Within Groups & 261233.1668 & 570 & 458.3038014 & & & \\ \hline
		Total &  635074.8716 & 572 & & & & \\ \hline
	\end{tabular}
	\end{center} 
	P more than .05, which means that there is a significant difference in radiance of P.oceanica at three different depth (0.5, 1.5 and 2.5). 
	\label{tab:8}
\end{table}

\begin{table}[htbp]
	\rowcolors{1}{Ivory}{GhostWhite}
	\caption{Summary of the \textit{ANOVA one-way} analysis: results of the \textit{single factor} (depth) testing of the radiance of\textbf{P.oceanica} at various depths: 		0.5, 1.5 and 2.5 meters}
	\begin{center}
	\begin{tabular}{|c|c|c|c|c|}
		\hline\hline
		\textbf{Groups} & \textbf{Count} & \textbf{Sum} & \textbf{Average} & \textbf{Variance} \\ \hline\hline
		Column 1 & 191 &	760.126692 & 3.9797209 & 18.6948046 \\ \hline
		Column 2 & 191 &	4527.214218 & 23.7026922  & 251.4863978 \\ \hline
		Column 3 & 191 &	12465.19972 & 65.2628257 & 1104.7302015 \\ \hline
	\end{tabular}
	\end{center}
	P$>$0.05, which means that there is a significant difference in radiance of P.oceanica at three different depth (0.5, 1.5 and 2.5). 
 	\label{tab:9}
\end{table}
\pagebreak

\rowcolors{1}{GhostWhite}{LightCyan}
\begin{longtable}{|c|c|c|c|c|c|c|}
	\caption{Results of the statistical analysis of spectral reflectance of \textbf{P.oceanica}, sets 1-350). Wavelength step: 3 nm. Measured on Agia Pelagia beach, 15th October}\label{tab:10} \\
	  \multicolumn{1}{|l|}{\textbf{wl}} &
	   \multicolumn{1}{l|}{\textbf{mean}} & 
	   \multicolumn{1}{l|}{\textbf{Q1}} & 
	   \multicolumn{1}{l|}{\textbf{min}} & 
	   \multicolumn{1}{l|}{\textbf{Q3}} & 
	   \multicolumn{1}{l|}{\textbf{max}} & 
	   \multicolumn{1}{l|}{\textbf{median}} \\ \hline
	\endfirsthead

@@

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{GNU-Irr-Splines.jpg}
		\caption{Irradiance of the seawater, measured in aquarium tank. Smooth Splines interpolation. Visualization in GNUplot­}
		\label{fig:48}
	\end{center}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.25]{GNU-Irr.jpg}
	\caption{Irradiance of the seawater, measured in aquarium tank. Smooth Bezier interpolation. Visualization in GNUplot}
	\label{fig:49}
\end{figure}
\pagebreak
\newpage
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.4]{G-gamma-rad.jpg}
		\caption{Frequency normality test against gamma distribution: radiance of the seawater, measured in aquarium tank. Visualization in Gretl.­}
		\label{fig:50}
	\end{center}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.4]{G-normal-dist.jpg}
	\caption{Frequency normality test against normal distribution: radiance of the seawater, measured in aquarium tank. Visualization in Gretl.­}
	\label{fig:51}
\end{figure}

@@

%%% Gretl start
\begin{table}[H]
	\rowcolors{1}{Thistle}{MintCream}
	\begin{center}
		\caption{Summary statistics of measurements set: 301-325. \\ Example of selected variables 12-22. Gretl.­}\label{tab:16}
		Summary Statistics, using the observations 301--325\\

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.38]{GNU-15.jpg}
		\caption{Multiplot of spectral reflectance of P.oceanica. Series 151-200. Shown midspread of the statistical quartiles Q1 and Q3 (vertical dashes) and mean value within the range (red bold dots). Gnuplot visualization­}
		\label{fig:59}
	\end{center}
\end{figure}
\pagebreak
